---
title: "Whale Shark Tourism Compliance Analysis"
author: "Hoa T.T. Ninh, Curtice R. Griffin, Rafael de la Para-Venegas, Kimberly L. Ovitz, Andy J. Danylchuk, Lucas P. Griffin"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    css: usf_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
if (!dir.exists("outputs")) dir.create("outputs")
```

# Load Packages

```{r packages}
suppressMessages({
  library(knitr)
  library(mgcv)
  library(mgcViz)
  library(dplyr)
  library(ggplot2)
  library(cowplot)
  library(lubridate)
  library(viridis)
})
```

# Data Loading and Preprocessing

```{r data-prep}
# Load data
general22 = read.csv("whale_shark_data/General2022.csv")
general16 = read.csv("whale_shark_data/General2016.csv")
boat22 = read.csv("whale_shark_data/Boat2022.csv")
boat16 = read.csv("whale_shark_data/Boat2016.csv")
swim22 = read.csv("whale_shark_data/Swimmer2022.csv")
swim16 = read.csv("whale_shark_data/Swimmer2016.csv")
infield22 = read.csv("whale_shark_data/Infield_2022.csv")
infield16_boat = read.csv("whale_shark_data/Infield2016_no_boats.csv")

# Load frequency data for violation and modeling analysis
encount_boat_2year_fre <- read.csv("whale_shark_data/encount_boat_2year_fre.csv")
encount_swim_2year_fre <- read.csv("whale_shark_data/encount_swim_2year_fre.csv")

# Adjust boat counts for police/survey boat presence
adjust_boat_count <- function(df) {
  df$NO_BOAT_AD <- ifelse(df$POLICE_PRESENCE == "yes" & df$SURVEY_BOAT == "yes",
                         df$NO_BOAT - 2,
                         ifelse(df$POLICE_PRESENCE == "yes" | df$SURVEY_BOAT == "yes",
                                df$NO_BOAT - 1,
                                df$NO_BOAT))
  return(df)
}

general22 <- adjust_boat_count(general22)
general16 <- adjust_boat_count(general16)

# Remove incomplete cases and outliers
clean_data <- function(df) {
  df <- df[complete.cases(df), ]
  df <- df %>% filter(FOOTAGE_AREA < 100000 & FOOTAGE_AREA > 10000)
  return(df)
}

general22 <- clean_data(general22)
general16 <- clean_data(general16)

# Clean boat and swimmer data
boat22 <- boat22[complete.cases(boat22), ]
boat16 <- boat16[complete.cases(boat16), ]
swim22 <- swim22[complete.cases(swim22), ]
swim16 <- swim16[complete.cases(swim16), ]

# Convert time formats
convert_time_format <- function(time_int) {
  time_str <- sprintf("%04d", time_int)
  minutes <- substr(time_str, 1, 2)
  seconds <- substr(time_str, 3, 4)
  paste0(minutes, ":", seconds)
}

# Apply time conversion to datasets that have V_LENGTH column
general22$V_LENGTH <- sapply(general22$V_LENGTH, convert_time_format)
general16$V_LENGTH <- sapply(general16$V_LENGTH, convert_time_format)
```

# Study Effort

```{r flight-duration}
# Calculate flight durations
duration_2016 <- general16 %>%
  mutate(V_LENGTH_SECOND = as.numeric(lubridate::ms(V_LENGTH))) %>%
  group_by(FLIGHT) %>%
  summarise(
    FLIGHT = first(FLIGHT),
    V_LENGTH_SECOND = first(V_LENGTH_SECOND),
    .groups = 'drop'
  ) %>%
  group_by(FLIGHT) %>%
  summarise(
    V_LENGTH_SECOND_FLIGHT = sum(V_LENGTH_SECOND),
    V_LENGTH_MIN_FLIGHT = sum(V_LENGTH_SECOND)/60,
    .groups = 'drop'
  )

duration_2022 <- general22 %>%
  mutate(V_LENGTH_SECOND = as.numeric(lubridate::ms(V_LENGTH))) %>%
  group_by(FLIGHT_C) %>%
  summarise(
    FLIGHT = first(FLIGHT),
    V_LENGTH_SECOND = first(V_LENGTH_SECOND),
    .groups = 'drop'
  ) %>%
  group_by(FLIGHT) %>%
  summarise(
    V_LENGTH_SECOND_FLIGHT = sum(V_LENGTH_SECOND),
    V_LENGTH_MIN_FLIGHT = sum(V_LENGTH_SECOND)/60,
    .groups = 'drop'
  )

# Video frame and flight statistics
video_frame_stats <- data.frame(
  Year = c("2016", "2022"),
  Video_Frames = c(nrow(general16), nrow(general22)),
  Flights = c(nrow(duration_2016), nrow(duration_2022)),
  Total_Hours = c(sum(duration_2016$V_LENGTH_MIN_FLIGHT)/60, sum(duration_2022$V_LENGTH_MIN_FLIGHT)/60),
  Mean_Duration = c(mean(duration_2016$V_LENGTH_MIN_FLIGHT), mean(duration_2022$V_LENGTH_MIN_FLIGHT)),
  SD_Duration = c(sd(duration_2016$V_LENGTH_MIN_FLIGHT), sd(duration_2022$V_LENGTH_MIN_FLIGHT))
)

knitr::kable(video_frame_stats, digits = 1, caption = "Video Frame and Flight Effort Summary")
```

# Density and Swimmer to Whale Shark Ratio Analysis

```{r density-calculations}
# Calculate densities per hectare
calculate_densities <- function(df) {
  df %>%
    mutate(
      WS_FOOTAGE = NO_WS/(FOOTAGE_AREA*0.0001),
      BOAT_FOOTAGE = NO_BOAT_AD/(FOOTAGE_AREA*0.0001),
      SWIMMER_FOOTAGE = NO_SWIMMER/(FOOTAGE_AREA*0.0001)
    )
}

general16 <- calculate_densities(general16)
general22 <- calculate_densities(general22)

# Calculate swimmers per whale shark from whale shark encounters
WS_2016_swim = swim16 %>%
  group_by(WS_CODE) %>%
  summarise(SUM_SWIMMERS_APR = sum(G_NO_SWIM, na.rm = TRUE), .groups = 'drop')

WS_2022_swim = swim22 %>%
  group_by(WS_CODE) %>%
  summarise(SUM_SWIMMERS_APR = sum(G_NO_SWIM, na.rm = TRUE), .groups = 'drop')

# Summary statistics and tests
density_metrics <- c("WS_FOOTAGE", "BOAT_FOOTAGE", "SWIMMER_FOOTAGE")
metric_names <- c("Whale sharks per hectare", "Boats per hectare", "Swimmers per hectare")

density_results <- data.frame(
  Metric = metric_names,
  Year_2016 = sapply(density_metrics, function(x) paste0(round(mean(general16[[x]]), 1), " ± ", round(sd(general16[[x]]), 1))),
  Year_2022 = sapply(density_metrics, function(x) paste0(round(mean(general22[[x]]), 1), " ± ", round(sd(general22[[x]]), 1))),
  W_Statistic = sapply(density_metrics, function(x) wilcox.test(general16[[x]], general22[[x]])$statistic),
  P_Value = sapply(density_metrics, function(x) {
    p <- wilcox.test(general16[[x]], general22[[x]])$p.value
    format(p, scientific = TRUE, digits = 3)
  })
)

# Add swimmers per whale shark results
swimmers_ws_test <- wilcox.test(WS_2016_swim$SUM_SWIMMERS_APR, WS_2022_swim$SUM_SWIMMERS_APR)
swimmers_ws_p <- format(swimmers_ws_test$p.value, scientific = TRUE, digits = 3)

# Combine all results
all_results <- rbind(
  density_results,
  data.frame(
    Metric = "Swimmers per whale shark",
    Year_2016 = paste0(round(mean(WS_2016_swim$SUM_SWIMMERS_APR), 1), " ± ", round(sd(WS_2016_swim$SUM_SWIMMERS_APR), 1)),
    Year_2022 = paste0(round(mean(WS_2022_swim$SUM_SWIMMERS_APR), 1), " ± ", round(sd(WS_2022_swim$SUM_SWIMMERS_APR), 1)),
    W_Statistic = swimmers_ws_test$statistic,
    P_Value = swimmers_ws_p
  )
)

knitr::kable(all_results, caption = "Density and Violation Comparisons Between Years", row.names = FALSE)
```

# Logbook Boat Counts

```{r logbook-boats}
# Boat counts from logbook data
logbook_2016 <- paste0(round(mean(infield16_boat$NO_BOAT), 1), " ± ", round(sd(infield16_boat$NO_BOAT), 1))
logbook_2022 <- paste0(round(mean(infield22$Boat_total), 1), " ± ", round(sd(infield22$Boat_total), 1))

# Statistical test
logbook_test <- wilcox.test(infield16_boat$NO_BOAT, infield22$Boat_total)

logbook_results <- data.frame(
  Metric = "Logbook boat counts",
  Year_2016 = logbook_2016,
  Year_2022 = logbook_2022,
  Max_2016 = max(infield16_boat$NO_BOAT),
  Max_2022 = max(infield22$Boat_total),
  W_Statistic = logbook_test$statistic,
  P_Value = format(logbook_test$p.value, scientific = TRUE, digits = 3)
)

knitr::kable(logbook_results, caption = "Logbook Boat Count Comparisons", row.names = FALSE)
```

# Violation Patterns

## Boat Distance Violations

```{r boat-distance-violations}
# Filter distance violations
boat22_dis <- boat22[boat22$BOAT_VIOLATE != "no violate min dis", ]
boat16_dis <- boat16[boat16$BOAT_VIOLATE != "no violate min dis", ]

# Violation counts
violation_summary <- data.frame(
  Year = c("2016", "2022"),
  Total_Violations = c(nrow(boat16_dis), nrow(boat22_dis)),
  Ratio = c(round(nrow(boat16_dis)/nrow(boat22_dis), 1), 1)
)

knitr::kable(violation_summary, caption = "Boat Distance Violation Counts", row.names = FALSE)
```

## Swimmer Distance Violations

```{r swimmer-violations}
# Filter swimmer distance violations
WS_2022_swim_dist = swim22 %>% filter(SWIM_VIOLATION != "no violate")
WS_2016_swim_dist = swim16 %>% filter(SWIM_VIOLATION != "no violate")

# Touching violations
touching_2022 = sum(swim22$NO_TOUCHING, na.rm = TRUE)
touching_2016 = sum(swim16$NO_TOUCHING, na.rm = TRUE)

# Summary table
swimmer_violation_summary <- data.frame(
  Violation_Type = c("Distance violations (< 5m)", "Touching violations"),
  Year_2016 = c(nrow(WS_2016_swim_dist), touching_2016),
  Year_2022 = c(nrow(WS_2022_swim_dist), touching_2022),
  Ratio_2022_to_2016 = c(
    round(nrow(WS_2022_swim_dist)/nrow(WS_2016_swim_dist), 1),
    round(touching_2022/touching_2016, 1)
  )
)

knitr::kable(swimmer_violation_summary, caption = "Swimmer Violation Counts", row.names = FALSE)

```

## Multiple Boats per Whale Shark

```{r boat-violations}
# Filter violation data
boat22_viol <- boat22 %>% filter(BOAT_VIOLATE != "no violate min dis")
boat16_viol <- boat16 %>% filter(BOAT_VIOLATE != "no violate min dis")

# Calculate boats per whale shark event
analyze_boat_violations <- function(boat_data) {
  events_total <- boat_data %>% distinct(EVE) %>% nrow()
  
  boats_per_ws <- boat_data %>%
    group_by(WS_CODE) %>%
    summarise(EVE = first(EVE), boats_count = n(), .groups = 'drop')
  
  multi_boat_events <- boats_per_ws %>%
    filter(boats_count > 1) %>%
    distinct(EVE) %>%
    nrow()
  
  return(list(
    total_events = events_total,
    multi_boat_events = multi_boat_events,
    percentage = round((multi_boat_events/events_total)*100, 1)
  ))
}

boat_analysis_2022 <- analyze_boat_violations(boat22_viol)
boat_analysis_2016 <- analyze_boat_violations(boat16_viol)

multi_boat_summary <- data.frame(
  Year = c("2016", "2022"),
  Multi_Boat_Events = c(boat_analysis_2016$multi_boat_events, boat_analysis_2022$multi_boat_events),
  Total_Events = c(boat_analysis_2016$total_events, boat_analysis_2022$total_events),
  Percentage = c(boat_analysis_2016$percentage, boat_analysis_2022$percentage)
)

knitr::kable(multi_boat_summary, caption = "Multiple Boats per Whale Shark Events", row.names = FALSE)
```

## Multiple Swimmers per Whale Shark

```{r swimmer-violation-analysis}
# Filter swimmer violations
swim22_viol <- swim22 %>% filter(SWIM_VIOLATION != "no violate")
swim16_viol <- swim16 %>% filter(SWIM_VIOLATION != "no violate")

# Calculate swimmers per whale shark event
analyze_swimmer_violations <- function(swim_data) {
  events_total <- swim_data %>% distinct(EVE) %>% nrow()
  
  swimmers_per_ws <- swim_data %>%
    group_by(WS_CODE) %>%
    summarise(EVE = first(EVE), swimmers_count = sum(G_NO_SWIM, na.rm = TRUE), .groups = 'drop')
  
  multi_swimmer_events <- swimmers_per_ws %>%
    filter(swimmers_count > 3) %>%
    distinct(EVE) %>%
    nrow()
  
  return(list(
    total_events = events_total,
    multi_swimmer_events = multi_swimmer_events,
    percentage = round((multi_swimmer_events/events_total)*100, 1),
    max_swimmers = max(swimmers_per_ws$swimmers_count),
    mean_swimmers = mean(swimmers_per_ws$swimmers_count),
    sd_swimmers = sd(swimmers_per_ws$swimmers_count)
  ))
}

swimmer_analysis_2022 <- analyze_swimmer_violations(swim22_viol)
swimmer_analysis_2016 <- analyze_swimmer_violations(swim16_viol)

multi_swimmer_summary <- data.frame(
  Year = c("2016", "2022"),
  Multi_Swimmer_Events = c(swimmer_analysis_2016$multi_swimmer_events, swimmer_analysis_2022$multi_swimmer_events),
  Total_Events = c(swimmer_analysis_2016$total_events, swimmer_analysis_2022$total_events),
  Percentage = c(swimmer_analysis_2016$percentage, swimmer_analysis_2022$percentage),
  Max_Swimmers = c(swimmer_analysis_2016$max_swimmers, swimmer_analysis_2022$max_swimmers)
)

knitr::kable(multi_swimmer_summary, caption = "Multiple Swimmers per Whale Shark Events", row.names = FALSE)

# Swimmers to whale shark ratio from general footage
general22_SW_WS <- general22 %>% 
  mutate(SWIM_WS = NO_SWIMMER/NO_WS)

general16_SW_WS <- general16 %>% 
  mutate(SWIM_WS = NO_SWIMMER/NO_WS)

# Statistical test
swimmers_ratio_test <- wilcox.test(general16_SW_WS$SWIM_WS, general22_SW_WS$SWIM_WS)

swimmers_ratio_results <- data.frame(
  Metric = "Swimmers to whale shark ratio",
  Year_2016 = paste0(round(mean(general16_SW_WS$SWIM_WS), 1), " ± ", round(sd(general16_SW_WS$SWIM_WS), 1)),
  Year_2022 = paste0(round(mean(general22_SW_WS$SWIM_WS), 1), " ± ", round(sd(general22_SW_WS$SWIM_WS), 1)),
  W_Statistic = swimmers_ratio_test$statistic,
  P_Value = format(swimmers_ratio_test$p.value, scientific = TRUE, digits = 3),
  Fold_Increase = paste0(round(mean(general22_SW_WS$SWIM_WS)/mean(general16_SW_WS$SWIM_WS), 1), "x")
)

knitr::kable(swimmers_ratio_results, caption = "Swimmers to Whale Shark Ratio Analysis", row.names = FALSE)
```

# Violation Frequency Analysis

```{r violation-frequency}
# Filter by year
boat_freq_2016 <- encount_boat_2year_fre %>% filter(YEAR == "2016")
boat_freq_2022 <- encount_boat_2year_fre %>% filter(YEAR == "2022")
swim_freq_2016 <- encount_swim_2year_fre %>% filter(YEAR == "2016")
swim_freq_2022 <- encount_swim_2year_fre %>% filter(YEAR == "2022")

# Statistical tests
boat_freq_test <- wilcox.test(boat_freq_2016$FRE_BOAT_VIOLATION_min, boat_freq_2022$FRE_BOAT_VIOLATION_min)
swim_freq_test <- wilcox.test(swim_freq_2016$FRE_SWIM_VIOLATION_min, swim_freq_2022$FRE_SWIM_VIOLATION_min)
touch_freq_test <- wilcox.test(swim_freq_2016$FRE_TOUCH_min, swim_freq_2022$FRE_TOUCH_min)

# Frequency analysis results
frequency_results <- data.frame(
  Violation_Type = c("Boat distance violations", "Swimmer distance violations (max 24 < 5m)", "Touching violations"),
  Year_2016 = c(
    paste0(round(mean(boat_freq_2016$FRE_BOAT_VIOLATION_min), 1), " ± ", round(sd(boat_freq_2016$FRE_BOAT_VIOLATION_min), 1)),
    paste0(round(mean(swim_freq_2016$FRE_SWIM_VIOLATION_min), 1), " ± ", round(sd(swim_freq_2016$FRE_SWIM_VIOLATION_min), 1)),
    paste0(round(mean(swim_freq_2016$FRE_TOUCH_min), 1), " ± ", round(sd(swim_freq_2016$FRE_TOUCH_min), 1))
  ),
  Year_2022 = c(
    paste0(round(mean(boat_freq_2022$FRE_BOAT_VIOLATION_min), 1), " ± ", round(sd(boat_freq_2022$FRE_BOAT_VIOLATION_min), 1)),
    paste0(round(mean(swim_freq_2022$FRE_SWIM_VIOLATION_min), 1), " ± ", round(sd(swim_freq_2022$FRE_SWIM_VIOLATION_min), 1)),
    paste0(round(mean(swim_freq_2022$FRE_TOUCH_min), 1), " ± ", round(sd(swim_freq_2022$FRE_TOUCH_min), 1))
  ),
  W_Statistic = c(boat_freq_test$statistic, swim_freq_test$statistic, touch_freq_test$statistic),
  P_Value = c(
    format(boat_freq_test$p.value, scientific = TRUE, digits = 3),
    format(swim_freq_test$p.value, scientific = TRUE, digits = 3),
    format(touch_freq_test$p.value, scientific = TRUE, digits = 3)
  )
)

knitr::kable(frequency_results, caption = "Violation Frequency per Minute Analysis", row.names = FALSE)
```

# Patrol Boat Effectiveness

```{r patrol-analysis}
# Patrol boat visibility analysis
patrol_presence_general <- general22 %>% 
  left_join(infield22, by = "FLIGHT") %>%
  filter(POLICE_PRESENCE == "yes")

video_frame_visibility <- data.frame(
  Metric = "Patrol boat visible in video frames",
  Visible_Frames = nrow(patrol_presence_general),
  Total_Frames = nrow(general22),
  Percentage = round((nrow(patrol_presence_general)/nrow(general22))*100, 1)
)

knitr::kable(video_frame_visibility, caption = "Patrol Boat Video Frame Visibility", row.names = FALSE)

# Patrol effectiveness analysis
patrol_freq_2022 <- encount_boat_2year_fre %>% 
  filter(YEAR == "2022") %>%
  left_join(infield22, by = "FLIGHT")

patrol_swim_freq_2022 <- encount_swim_2year_fre %>% 
  filter(YEAR == "2022") %>%
  left_join(infield22, by = "FLIGHT")

# Filter by patrol presence
boat_patrol_present <- patrol_freq_2022 %>% filter(POLICE_INFIELD == "Y")
boat_patrol_absent <- patrol_freq_2022 %>% filter(POLICE_INFIELD == "N")
swim_patrol_present <- patrol_swim_freq_2022 %>% filter(POLICE_INFIELD == "Y")
swim_patrol_absent <- patrol_swim_freq_2022 %>% filter(POLICE_INFIELD == "N")

# Statistical tests for patrol effectiveness
boat_patrol_test <- wilcox.test(boat_patrol_present$FRE_BOAT_VIOLATION_min, boat_patrol_absent$FRE_BOAT_VIOLATION_min)
swim_patrol_test <- wilcox.test(swim_patrol_present$FRE_SWIM_VIOLATION_min, swim_patrol_absent$FRE_SWIM_VIOLATION_min)
touch_patrol_test <- wilcox.test(swim_patrol_present$FRE_TOUCH_min, swim_patrol_absent$FRE_TOUCH_min)

# Patrol effectiveness results
patrol_effectiveness <- data.frame(
  Violation_Type = c("Boat distance violations", "Swimmer distance violations", "Touching violations"),
  Patrol_Present = c(
    paste0(round(mean(boat_patrol_present$FRE_BOAT_VIOLATION_min), 1), " ± ", round(sd(boat_patrol_present$FRE_BOAT_VIOLATION_min), 1)),
    paste0(round(mean(swim_patrol_present$FRE_SWIM_VIOLATION_min), 1), " ± ", round(sd(swim_patrol_present$FRE_SWIM_VIOLATION_min), 1)),
    paste0(round(mean(swim_patrol_present$FRE_TOUCH_min), 1), " ± ", round(sd(swim_patrol_present$FRE_TOUCH_min), 1))
  ),
  Patrol_Absent = c(
    paste0(round(mean(boat_patrol_absent$FRE_BOAT_VIOLATION_min), 1), " ± ", round(sd(boat_patrol_absent$FRE_BOAT_VIOLATION_min), 1)),
    paste0(round(mean(swim_patrol_absent$FRE_SWIM_VIOLATION_min), 1), " ± ", round(sd(swim_patrol_absent$FRE_SWIM_VIOLATION_min), 1)),
    paste0(round(mean(swim_patrol_absent$FRE_TOUCH_min), 1), " ± ", round(sd(swim_patrol_absent$FRE_TOUCH_min), 1))
  ),
  W_Statistic = c(boat_patrol_test$statistic, swim_patrol_test$statistic, touch_patrol_test$statistic),
  P_Value = c(
    format(boat_patrol_test$p.value, scientific = TRUE, digits = 3),
    format(swim_patrol_test$p.value, scientific = TRUE, digits = 3),
    format(touch_patrol_test$p.value, scientific = TRUE, digits = 3)
  )
)

knitr::kable(patrol_effectiveness, caption = "Patrol Boat Effectiveness Analysis", row.names = FALSE)
```

# Predictive Modeling

```{r modeling}
# Prepare data
encount_boat_2year_fre$DATE <- as.factor(encount_boat_2year_fre$DATE)
encount_swim_2year_fre$DATE <- as.factor(encount_swim_2year_fre$DATE)

## Boat Violation Models

# Check predictor correlations
cor_matrix <- cor(encount_boat_2year_fre[, c("SUM_WS_FLIGHT_min", "SUM_SWIM_FLIGHT_min", "SUM_BOAT_FLIGHT_min", "RATIO_WS_SWIM_min")])
knitr::kable(round(cor_matrix, 3), caption = "Boat Model Predictor Correlations")

# Model 1: All variables with interaction (te) and one smooth (s)
model_1 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 te(SUM_WS_FLIGHT_min, SUM_BOAT_FLIGHT_min, k = c(3, 4)) + 
                 s(RATIO_WS_SWIM_min) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 2: Interaction (te) without one smooth (s)
model_2 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 te(SUM_WS_FLIGHT_min, SUM_BOAT_FLIGHT_min, k = c(3, 4)) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 3: All variables as separate smooths
model_3 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_WS_FLIGHT_min, k = 3) + 
                 s(SUM_BOAT_FLIGHT_min, k = 4) + 
                 s(RATIO_WS_SWIM_min) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 4: Two smooths without interaction (WS and Boats)
model_4 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_WS_FLIGHT_min, k = 3) + 
                 s(SUM_BOAT_FLIGHT_min, k = 4) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 5: Two smooths without interaction (WS and Ratio)
model_5 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_WS_FLIGHT_min, k = 3) + 
                 s(RATIO_WS_SWIM_min) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 6: Two smooths without interaction (Boats and Ratio)
model_6 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_BOAT_FLIGHT_min, k = 4) + 
                 s(RATIO_WS_SWIM_min) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 7: One smooth without interaction (WS only)
model_7 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_WS_FLIGHT_min, k = 3) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 8: One smooth without interaction (Boats only)
model_8 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(SUM_BOAT_FLIGHT_min, k = 4) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 9: One smooth without interaction (Ratio only)
model_9 <- gam(FRE_BOAT_VIOLATION_min ~ 
                 s(RATIO_WS_SWIM_min) + 
                 s(DATE, bs = "re"), 
               data = encount_boat_2year_fre,
               family = Gamma(link = "log"),
               method = "REML")

# Model 10: Only random effect
model_10 <- gam(FRE_BOAT_VIOLATION_min ~ 
                  s(DATE, bs = "re"), 
                data = encount_boat_2year_fre,
                family = Gamma(link = "log"),
                method = "REML")

# List models and create comparison table
boat_models <- list(model_1, model_2, model_3, model_4, model_5, model_6, model_7, model_8, model_9, model_10)

# Function to check if term is present in model formula
check_term <- function(model, term_pattern) {
  formula_str <- as.character(formula(model))[3]
  ifelse(grepl(term_pattern, formula_str), "X", "")
}

# Create data-driven model comparison table
boat_model_info <- data.frame(
  Model = paste0("Model_", 1:10),
  "No_of_WS" = sapply(boat_models, function(x) check_term(x, "SUM_WS_FLIGHT_min")),
  "Ratio_swimmers_per_WS" = sapply(boat_models, function(x) check_term(x, "RATIO_WS_SWIM_min")),
  "No_of_boats" = sapply(boat_models, function(x) check_term(x, "SUM_BOAT_FLIGHT_min")),
  "WS_boats_Interaction" = sapply(boat_models, function(x) check_term(x, "te\\(SUM_WS_FLIGHT_min, SUM_BOAT_FLIGHT_min")),
  "Sampling_day" = sapply(boat_models, function(x) check_term(x, "s\\(DATE")),
  AIC = round(sapply(boat_models, AIC), digits = 2)
)

boat_model_info <- boat_model_info %>% arrange(AIC)
knitr::kable(boat_model_info, caption = "Boat Violation Model Selection (Ordered by AIC)")

# Export to CSV
write.csv(boat_model_info, "outputs/model_selection_table_boat.csv", row.names = FALSE)

# Selected boat model
m_gam_boat <- model_2
```

```{r boat-diagnostics, fig.width=12, fig.height=8}
# Model diagnostics
plot(m_gam_boat, pages = 1, residuals = TRUE, rug = TRUE)
gam.check(m_gam_boat)

# ACF plot
acf(resid(m_gam_boat), lag.max = 36, main = "ACF - Boat Model")

# Fitted vs residuals
fitted_values <- fitted(m_gam_boat)
residuals <- resid(m_gam_boat)

plot(fitted_values, residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Fitted Values vs Residuals - Boat Model",
     pch = 16, col = "blue", cex = 0.7)
abline(h = 0, col = "red", lty = 2, lwd = 2)

# mgcViz diagnostics
viz_boat <- getViz(m_gam_boat)
check(viz_boat,
      a.qq = list(method = "tnorm", a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))
```

```{r boat-interaction-plot}
# Create interaction plot for boat model
SUM_WS_FLIGHT_min_seq <- seq(min(encount_boat_2year_fre$SUM_WS_FLIGHT_min), 
                             max(encount_boat_2year_fre$SUM_WS_FLIGHT_min), 
                             length.out = 100)
SUM_BOAT_FLIGHT_min_seq <- seq(min(encount_boat_2year_fre$SUM_BOAT_FLIGHT_min), 
                               max(encount_boat_2year_fre$SUM_BOAT_FLIGHT_min), 
                               length.out = 100)

grid_boat <- expand.grid(SUM_WS_FLIGHT_min = SUM_WS_FLIGHT_min_seq,
                        SUM_BOAT_FLIGHT_min = SUM_BOAT_FLIGHT_min_seq,
                        RATIO_WS_SWIM_min = mean(encount_boat_2year_fre$RATIO_WS_SWIM_min),
                        DATE = factor(levels(encount_boat_2year_fre$DATE)[1]))

pred_boat <- predict.gam(m_gam_boat, newdata = grid_boat, type = "link", se.fit = TRUE, exclude = "s(DATE)")

grid_boat$fit_link <- pred_boat$fit
grid_boat$se.fit <- pred_boat$se.fit
grid_boat$lower_link <- grid_boat$fit_link - 1.96 * grid_boat$se.fit
grid_boat$upper_link <- grid_boat$fit_link + 1.96 * grid_boat$se.fit

inverse_link <- family(m_gam_boat)$linkinv
grid_boat$fit_response <- inverse_link(grid_boat$fit_link)
grid_boat$lower_response <- inverse_link(grid_boat$lower_link)
grid_boat$upper_response <- inverse_link(grid_boat$upper_link)

plot_interaction_boat <- ggplot(grid_boat, aes(x = SUM_WS_FLIGHT_min, y = SUM_BOAT_FLIGHT_min, fill = fit_response)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted number of boat \nviolations ") +
  labs(x = "No. of WS", y = "No. of boats") +
  theme_classic() +
  theme(legend.position = "top",
        text = element_text(size = 13),       
        axis.title = element_text(size = 13), 
        axis.text = element_text(size = 13))

print(plot_interaction_boat)
```

## Swimmer Violation Models

```{r swimmer-models}
# Check predictor correlations
cor_matrix_swim <- cor(encount_swim_2year_fre[, c("SUM_WS_FLIGHT_min", "SUM_SWIM_FLIGHT_min", "SUM_BOAT_FLIGHT_min", "RATIO_WS_SWIM_min")])
knitr::kable(round(cor_matrix_swim, 3), caption = "Swimmer Model Predictor Correlations")

# Model 1: All variables with interaction (te) and one smooth (s)
model_swim_1 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     te(SUM_WS_FLIGHT_min, SUM_SWIM_FLIGHT_min, k = c(4, 4)) + 
                     s(RATIO_WS_SWIM_min) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 2: Interaction (te) without one smooth (s)
model_swim_2 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     te(SUM_WS_FLIGHT_min, SUM_SWIM_FLIGHT_min, k = c(4, 4)) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 3: All variables as separate smooths
model_swim_3 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_WS_FLIGHT_min, k = 4) + 
                     s(SUM_SWIM_FLIGHT_min, k = 4) + 
                     s(RATIO_WS_SWIM_min) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 4: Two smooths without interaction (WS and Swimmers)
model_swim_4 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_WS_FLIGHT_min, k = 4) + 
                     s(SUM_SWIM_FLIGHT_min, k = 4) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 5: Two smooths without interaction (WS and Ratio)
model_swim_5 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_WS_FLIGHT_min, k = 4) + 
                     s(RATIO_WS_SWIM_min) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 6: Two smooths without interaction (Swimmers and Ratio)
model_swim_6 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_SWIM_FLIGHT_min, k = 4) + 
                     s(RATIO_WS_SWIM_min) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 7: One smooth without interaction (WS only)
model_swim_7 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_WS_FLIGHT_min, k = 4) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 8: One smooth without interaction (Swimmers only)
model_swim_8 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(SUM_SWIM_FLIGHT_min, k = 4) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 9: One smooth without interaction (Ratio only)
model_swim_9 <- gam(FRE_SWIM_VIOLATION_min ~ 
                     s(RATIO_WS_SWIM_min) + 
                     s(DATE, bs = "re"), 
                   data = encount_swim_2year_fre,
                   family = Gamma(link = "log"),
                   method = "REML")

# Model 10: Only random effect
model_swim_10 <- gam(FRE_SWIM_VIOLATION_min ~ 
                      s(DATE, bs = "re"), 
                    data = encount_swim_2year_fre,
                    family = Gamma(link = "log"),
                    method = "REML")

# List models and create comparison table
swim_models <- list(model_swim_1, model_swim_2, model_swim_3, model_swim_4, model_swim_5, 
                   model_swim_6, model_swim_7, model_swim_8, model_swim_9, model_swim_10)

# Create data-driven swimmer model comparison table
swim_model_info <- data.frame(
  Model = paste0("Model_", 1:10),
  "No_of_WS" = sapply(swim_models, function(x) check_term(x, "SUM_WS_FLIGHT_min")),
  "Ratio_swimmers_per_WS" = sapply(swim_models, function(x) check_term(x, "RATIO_WS_SWIM_min")),
  "No_of_swimmers" = sapply(swim_models, function(x) check_term(x, "SUM_SWIM_FLIGHT_min")),
  "WS_swimmers_Interaction" = sapply(swim_models, function(x) check_term(x, "te\\(SUM_WS_FLIGHT_min, SUM_SWIM_FLIGHT_min")),
  "Sampling_day" = sapply(swim_models, function(x) check_term(x, "s\\(DATE")),
  AIC = round(sapply(swim_models, AIC), digits = 2)
)

swim_model_info <- swim_model_info %>% arrange(AIC)
knitr::kable(swim_model_info, caption = "Swimmer Violation Model Selection (Ordered by AIC)")

# Export to CSV
write.csv(swim_model_info, "outputs/model_selection_table_swimmer.csv", row.names = FALSE)

# Selected swimmer model
m_gam_swim <- model_swim_1
```

```{r swimmer-diagnostics, fig.width=12, fig.height=8}
# Model diagnostics
plot(m_gam_swim, pages = 1, residuals = TRUE, rug = TRUE)
gam.check(m_gam_swim)

# ACF plot
acf(resid(m_gam_swim), main = "ACF - Swimmer Model")

# Fitted vs residuals
fitted_values_swim <- fitted(m_gam_swim)
residuals_swim <- resid(m_gam_swim)

plot(fitted_values_swim, residuals_swim,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Fitted Values vs Residuals - Swimmer Model",
     pch = 16, col = "blue", cex = 0.7)
abline(h = 0, col = "red", lty = 2, lwd = 2)

# mgcViz diagnostics
viz_swim <- getViz(m_gam_swim)
check(viz_swim,
      a.qq = list(method = "tnorm", a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))
```

```{r swimmer-plots}
# Create interaction plot for swimmer model
SUM_WS_FLIGHT_min_seq <- seq(min(encount_swim_2year_fre$SUM_WS_FLIGHT_min), 
                             max(encount_swim_2year_fre$SUM_WS_FLIGHT_min), 
                             length.out = 100)
SUM_SWIM_FLIGHT_min_seq <- seq(min(encount_swim_2year_fre$SUM_SWIM_FLIGHT_min), 
                               max(encount_swim_2year_fre$SUM_SWIM_FLIGHT_min), 
                               length.out = 100)

grid_swim <- expand.grid(SUM_WS_FLIGHT_min = SUM_WS_FLIGHT_min_seq,
                        SUM_SWIM_FLIGHT_min = SUM_SWIM_FLIGHT_min_seq,
                        RATIO_WS_SWIM_min = mean(encount_swim_2year_fre$RATIO_WS_SWIM_min),
                        DATE = factor(levels(encount_swim_2year_fre$DATE)[1]))

pred_swim <- predict.gam(m_gam_swim, newdata = grid_swim, type = "link", se.fit = TRUE, exclude = "s(DATE)")

grid_swim$fit_link <- pred_swim$fit
grid_swim$se.fit <- pred_swim$se.fit
grid_swim$lower_link <- grid_swim$fit_link - 1.96 * grid_swim$se.fit
grid_swim$upper_link <- grid_swim$fit_link + 1.96 * grid_swim$se.fit

inverse_link <- family(m_gam_swim)$linkinv
grid_swim$fit_response <- inverse_link(grid_swim$fit_link)
grid_swim$lower_response <- inverse_link(grid_swim$lower_link)
grid_swim$upper_response <- inverse_link(grid_swim$upper_link)

plot_interaction_swim <- ggplot(grid_swim, aes(x = SUM_WS_FLIGHT_min, y = SUM_SWIM_FLIGHT_min, fill = fit_response)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted number of swimmer \nviolations ") +
  labs(x = "No. of WS", y = "No. of swimmers") +
  theme_classic() +
  theme(legend.position = "top",
        text = element_text(size = 13),       
        axis.title = element_text(size = 13), 
        axis.text = element_text(size = 13))

print(plot_interaction_swim)

# Ratio effect plot
new_data_ratio <- data.frame(
  RATIO_WS_SWIM_min = seq(min(encount_swim_2year_fre$RATIO_WS_SWIM_min),
                          max(encount_swim_2year_fre$RATIO_WS_SWIM_min), length.out = 100),
  SUM_SWIM_FLIGHT_min = mean(encount_swim_2year_fre$SUM_SWIM_FLIGHT_min),
  SUM_BOAT_FLIGHT_min = mean(encount_swim_2year_fre$SUM_BOAT_FLIGHT_min),
  SUM_WS_FLIGHT_min = mean(encount_swim_2year_fre$SUM_WS_FLIGHT_min),
  DATE = factor(levels(encount_swim_2year_fre$DATE)[1], levels = levels(encount_swim_2year_fre$DATE))
)

pred_ratio <- predict.gam(m_gam_swim, newdata = new_data_ratio, type = "link", se.fit = TRUE, exclude = "s(DATE)")

new_data_ratio$fit_link <- pred_ratio$fit
new_data_ratio$se.fit <- pred_ratio$se.fit
new_data_ratio$lower_link <- new_data_ratio$fit_link - 1.96 * new_data_ratio$se.fit
new_data_ratio$upper_link <- new_data_ratio$fit_link + 1.96 * new_data_ratio$se.fit

inverse_link <- family(m_gam_swim)$linkinv
new_data_ratio$fit_response <- inverse_link(new_data_ratio$fit_link)
new_data_ratio$lower_response <- inverse_link(new_data_ratio$lower_link)
new_data_ratio$upper_response <- inverse_link(new_data_ratio$upper_link)

plot_ratio <- ggplot(new_data_ratio, aes(x = RATIO_WS_SWIM_min, y = fit_response)) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = lower_response, ymax = upper_response), fill = "black", alpha = 0.2) +
  labs(x = "Ratio of swimmers per WS", y = "Predicted number of swimmer violations ") +
  theme_bw() +  
  geom_point(data = encount_swim_2year_fre, aes(x = RATIO_WS_SWIM_min, y = FRE_SWIM_VIOLATION_min), color = "blue") +
  theme(text = element_text(size = 13),       
        axis.title = element_text(size = 13), 
        axis.text = element_text(size = 13))

print(plot_ratio)
```

```{r combined-plots}
# Combine plots
combined_plot <- cowplot::plot_grid(
  plot_interaction_boat, plot_interaction_swim,
  labels = "auto", 
  ncol = 2
) 

print(combined_plot)

# Save plot
ggsave("outputs/GAM_output.png", combined_plot, height = 7, width = 9)

# Model summaries
summary(m_gam_boat) 
summary(m_gam_swim) 
```

# Summary

```{r summary-table}
# Create summary of key findings
summary_findings <- data.frame(
  Metric = c(
    "Whale shark density (per hectare)",
    "Boat density (per hectare)", 
    "Swimmer density (per hectare)",
    "Swimmers per whale shark ratio",
    "Multiple boats per shark (%)",
    "4+ swimmers per shark (%)",
    "Patrol boat not visible (flights)"
  ),
  Year_2016 = c(
    paste0(round(mean(general16$WS_FOOTAGE), 1), " ± ", round(sd(general16$WS_FOOTAGE), 1)),
    paste0(round(mean(general16$BOAT_FOOTAGE), 1), " ± ", round(sd(general16$BOAT_FOOTAGE), 1)),
    paste0(round(mean(general16$SWIMMER_FOOTAGE), 1), " ± ", round(sd(general16$SWIMMER_FOOTAGE), 1)),
    paste0(round(mean(general16_SW_WS$SWIM_WS), 1), " ± ", round(sd(general16_SW_WS$SWIM_WS), 1)),
    paste0(boat_analysis_2016$percentage, "%"),
    paste0(swimmer_analysis_2016$percentage, "%"),
    "N/A"
  ),
  Year_2022 = c(
    paste0(round(mean(general22$WS_FOOTAGE), 1), " ± ", round(sd(general22$WS_FOOTAGE), 1)),
    paste0(round(mean(general22$BOAT_FOOTAGE), 1), " ± ", round(sd(general22$BOAT_FOOTAGE), 1)),
    paste0(round(mean(general22$SWIMMER_FOOTAGE), 1), " ± ", round(sd(general22$SWIMMER_FOOTAGE), 1)),
    paste0(round(mean(general22_SW_WS$SWIM_WS), 1), " ± ", round(sd(general22_SW_WS$SWIM_WS), 1)),
    paste0(boat_analysis_2022$percentage, "%"),
    paste0(swimmer_analysis_2022$percentage, "%"),
    paste0(patrol_absent_flights, "/", total_flights_2022)
  )
)

knitr::kable(summary_findings, caption = "Summary of Key Findings")

# Export results
write.csv(summary_findings, "outputs/whale_shark_analysis_summary.csv", row.names = FALSE)
write.csv(density_results, "outputs/density_analysis_results.csv", row.names = FALSE)
```

